Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-06 paper/paper.tex
added second picture
first text for result

    \newcommand{\param}[1]{\texttt{#1}}
    \newcommand{\feedback}{\param{feedback}}
    \newcommand{\gain}{\param{gain}}
    \newcommand{\size}{N}
    \newcommand{\ERprob}{p}
    \newcommand{\RINGk}{k}
    \newcommand{\FFlayers}{l}

@@

\paragraph{Genetic Optimization}
Due to the stochastic nature of the fitness function $F_C(G)$ of a genotype
$G$, a very high number of samples would be required to obtain sufficiently
smooth approximations of $F_C(G)$ for typical optimization methods like
gradient descend or simulated annealing. But as the computational cost for
even one learning process was considerable high, these methods were not
feasible.

%and the fact that single evaluations of $F_C$ are
%computationally very expensive, thus preventing increasing the number of
%samples such that the estimation $E(F_C)$ becomes sufficiently smooth,
%typical optimization methods like gradient descent or simulated annealing
%are not usable.

@@

\paragraph{Choice of Parameters}
- $N=100$ fixed, since greater $N$ always increased quality and computational cost
increased quadraticaly to $N$
- Mutation rate

\section{Results}
\subsection{Frequency Dependent Optimizations}
- short description: get feeling for frequency range (by using unoptimized parameters
from the force paper), make optimizations with 3 different challenges: 1) challenges that cover the frequency area completely, challenges that cover low frequencies, and challanges that cover high frequencies, then pick genotypes with high fitness.
\subsection{Topology Optimizations}
Baseline
gain and feedback tests

(feedback shifts possible frequencies to higher values)

success of omega: erdös-renyi, ring
optimizations: ER topology, Ring Topology, Mixing

\begin{figure}
    {\footnotesize\input{pictures/freqrange}}
    \caption{CAPTION TEXT}
    \label{fig:freqrange}
\end{figure}

@@

- Problems
    * Many magic parameters scattered in the model that were not examined exhaustively
    * Only few and uncomprehensive number of test / experiments conducted due to the
      computational cost.
    * Test with several other topologies did not yield concrete results,
      only rough tendencies.
    * Influence of statistics: Learning success depends stronger on choice
      of challenges than parameters (partially) -> must be very careful how to
      interpret the results

\blindtext

>>>>

    \newcommand{\param}[1]{\texttt{#1}}
    \newcommand{\feedback}{\param{feedback}}
    \newcommand{\gain}{\param{gain}}
    \newcommand{\size}{N}
    \newcommand{\ERprob}{p}
    \newcommand{\RINGk}{k}
    \newcommand{\ERtoRING}{r}
    \newcommand{\FFlayers}{l}

@@

\paragraph{Genetic Optimization}
Due to the stochastic nature of a genotype $G$'s fitness $F_C(G)$,
a very high number of samples would be required to obtain sufficiently
smooth approximations of $F_C(G)$ for typical optimization methods like
gradient descend or simulated annealing. But as the computational cost for
even one learning process was considerable high, these methods were not
feasible.

@@

\paragraph{Choice of Parameters and Algorithmic Quantities}
For all of the simulations we chose the size of the networks to be fixed
with $N=100$. Though increasing $N$ generally improved the quality of the
learning results, the computational cost increased proportional to $N^2$,
which made using networks as large as in the original FORCE publication
(between $500$ and $20000$) utterly unusable for our purposes.

The mutation rate of the parameters $\Theta$ of the genotypes\dots

\section{Results}
\subsection{Optimizing Genotypes For Frequency Ranges}
Since we wanted to examine how the ideal genotypes vary for different
types of problems, we examined how the choice of a frequency range affects
the optimized genotype parameters.
To this end we looked at sparse random networks (Erdös-Renyi topology with
connection probability $\ERprob$) and conducted optimizations of the
parameters $\gain$, $\feedback$ and $\ERprob$ for both high ($C\ix{high}$)
and low ($C\ix{low}$) frequency challenges (see FIG.
\ref{fig:freq_optimization}).

\begin{figure}[hbct]
    {\footnotesize\input{pictures/freq_optimization}}
    \caption{
        Genetic optimization process of genotypes with Erdös-Renyi topology
        ($N=100$ fixed) for the challenges $C\ix{high}$ and $C\ix{low}$.
        In each figure, one can see the $50$ genotypes of every Generation
        (single dots) together with the respective mean values (thick
        lines). Figures (a) to (c) depict the evolution of the three
        parameters ($\gain$, $\feedback$, and $\ERprob$) being optimized,
        while figure (d) shows the development of the fitness values.
    }
    \label{fig:freq_optimization}
\end{figure}

The most striking result is the strong growth of the feedback for
$C\ix{high}$ to values higher than 14, while the value for $C\ix{low}$
converges quickly to approximately 1 (FIG. \ref{fig:freq_optimization} (b)).
This behavior may indicate, that the feedback pathway must
overshadow the (maybe too slow) internal mechanics for high frequencies,
while it is the other way round for low frequencies.
Indeed, FIG. \ref{fig:freq_optimization} (a) shows that genotypes with
a lower gain value (on average) are more successful for high frequencies,
which seems to fit in: excessive recurrent dynamics with a long time
scale (compared to the period to be learned) may disturb the learning.
That the single genotypes remain quite scattered throughout the whole
optimization process, though, implies that the exact gain value is not
critical for the success of a genotype.

Another interesting aspect is the branching process of $p$ that takes place
for $C\ix{low}$ in FIG. \ref{fig:freq_optimization} (c). Albeit the branch
seems to be dying out towards the end, it suggests, that very high $p$
values don't ruin the networks capacity to learn low frequencies. This
behavior was observed neither for high frequencies nor for optimizations
over the full frequency regime.

But how well did the fitness actually improve during the optimization?
When looking at FIG. \ref{fig:freq_optimization} (d), one sees that the
fitness for $C\ix{high}$ increased during nearly the whole process, while
the algorithm failed to further optimize the genotypes' fitness for
$C\ix{low}$ after about 20 generations. A major reason for this is
probably, that the initial population for the $C\ix{low}$ optimization
already contained genotypes with parameters similar to the (at least
locally) optimal ones.

One can also look at the success of the optimization algorithm in a more
specific way: FIG. \ref{fig:freqrange} shows how well an unoptimized
genotype $G_0$ handles fixed-frequency-challenges when compared to the
genotypes $G\ix{high}$ and $G\ix{low}$ optimized for
$C\ix{high}$ and $C\ix{low}$. Here it becomes clear that $G\ix{high}$ can
indeed succeed at higher frequencies than $G_0$, but that it also fails at
learning lower frequencies -- which was expected. On the contrary, while
the high frequency capacities of $G\ix{low}$ have decreased when compared
to $G_0$, the low frequency capacities have only marginally improved. This
reinforces the impression that the optimization was unable to produce
genotypes that can cope with low frequencies.

\begin{figure}
    {\footnotesize\input{pictures/freqrange}}
    \caption{
        The covered frequency range for an unoptimized
        genotype $G_0$ as well as genotypes $G\ix{high}$ and $G\ix{low}$
        that are optimized for $C\ix{high}$ respectively $C\ix{low}$. Each
        of the three genotypes was tested against constant challenges
        $C_\omega$ that yielded simple sinusoidal waves with frequency
        $\omega$ for every seed value. The topology of all genotypes was
        a mere Erdös-Renyi topology.
        The parameters for the unoptimized genotype $G_0$ were taken to be
        default values from \cite{FORCE} (which also uses Erdös-Renyi
        topologies), while we chose genotypes $G\ix{high}$ and $G\ix{low}$
        that did particularly well when explicitly optimizing for high and
        low frequencies (see FIG. \ref{fig:freq_optimization} for more
        details).
    }
    \label{fig:freqrange}
\end{figure}

\subsection{Topology Optimizations}
To find out whether the topology can be used to improve the learning, we
performed an optimization in which the networks were an interpolation
between sparse random connections (Erdös-Renyi) and a ring topology where
the $\RINGk$ nearest neighbors are connected (for non integer $\RINGk$, the
fractional part determines the probability to use $[\RINGk]$ or $[\RINGk]+1$).
Since we are interested in the influence of the topology, the parameters
gain and feedback are kept constant during the optimization process,
leaving the connection probability for ER topology $\ERprob$, the ring
connection length $\RINGk$ and the interpolation value $\ERtoRING$ between
ER and ring as free parameters.

The results of the optimization are depicted in FIG. \ref{}. Unfortunately,
the improvement in fitness after the first few generations in extremely
low. The development of the ration parameter, however, unambiguously shows
that the selection process prefers generators that consist mostly of ring
topology with ring parameter $\RINGk \approx 1$. The ER connection probability
shows a very broad distribution, which is simply due to the low dependence
of the generated networks on $\ERprob$ when the usage of ER topology
$\ERtoRING$ approaches zero.

@@

- Problems
    * Many magic / ``empiric'' parameters scattered in the model that were not examined exhaustively
    * Only few and uncomprehensive number of test / experiments conducted due to the
      computational cost.
    * Test with several other topologies did not yield concrete results,
      only rough tendencies.
    * Influence of statistics: Learning success depends stronger on choice
      of challenges than parameters (partially) -> must be very careful how to
      interpret the results / should have used fixed array of challenges instead
      of random sampling
